A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-28 sections/algorithm.tex
Section 3

The algorithmic typing reduction rules, defined in Figure~\ref{fig:alg}, have
the form $\Gm \rto \Gm'$.
% The worklists $\Gm, \Gm'$ contain both variable declarations and judgments.
The reduction process treats the worklist as a stack.  In every step it pops
the first judgment from the worklist for processing and possibly pushes new
judgments onto the worklist.  The syntax $\Gm\redto \Gm'$ denotes multiple
reduction steps. If $\Gm\redto\nil$, we say that $\Gm$ reduces successfully;
this corresponds to a successful type checking.

Let us explain the rules in more detail:
Rules 1-3 pop a variable declaration; this garbage collects variables that are
out of scope for the remaining judgements.
All other rules concern a judgement at the front of the worklist. Logically we
can discern 6 group of rules.

@@

\paragraph{\bf 2. Existential decomposition.}
Rules 10 and 11 are algorithmic versions of Rule $\mathtt{{\le}{\to}}$; they
both partially instantiate $\al$ to function types.
The domain $\al[1]$ and range $\al[2]$ of the new function type are not determined:
they are fresh existential variables with the same scope as $\al$.
We remove replace $\al$ in the worklist with  $\al[1], \al[2]$.
To propagate the instantiation to the rest of the worklist and maintain well-formedness,
every reference to $\al$ is replaced by $\al[1] \to \al[2]$.
The \emph{occurs-check} condition prevents divergence as usual.
For example, without it $\al \le 1 \to \al$ would diverge.

@@

\paragraph{\bf 5. Inference judgments.}
Inference judgments behave differently compared with subtyping and checking judgments:
they \emph{return} a type instead of only accepting or rejecting.
For algorithmic system, where guessings are involved,
it may happen that the output type of an inference judgment refers to new existential variables,
such as Rule 25.
In comparison to Rule 8 and 9, where new variables are only refered by the sub-derivation,
Rule 25 introduces variables $\al, \bt$ that affect the remaining judgment chain.
This rule is carefully designed so that the output variables are bound by earlier declarations,
thus the well-formedness of the worklist is preserved,
and the garbage will be collected at the correct time.
By making use of the continuation-passing-style judgement chain,
inner judgments always shares context with its parent judgment,
therefore we achieve all the desired property in such a neat way.

@@

Rule 25 infers the type of a lambda expression by introducing $\al, \bt$
as the input and output types of the function, respectively.
After checking the body $e$ under the assumption $x:\al$,
the return type might reflect more information than simply $\al \to \bt$
through propagation when existential variables are solved or partially solved.
The variable scopes are maintained during the process:
the assumption of argument type ($x:\al$) is recycled after checking against the function body;
the declaration of the function type ($\al,\bt$) is only visible to
the checking and the remaining chain $[\al\to\bt/a]\jg$.
The recycling process of Rule 25 differs from DK's corresponding rule significantly,
and we further discuss the differences in Section~\ref{sec:discussion:scoping}.

@@

\paragraph{\bf 6. Application inferences judgements}
Finally, Rules 27-29 deal with application inference judgments.
Rules 27 and 28 behaves similarly to declarative rules $\mathtt{Decl\forall App}$ and $\mathtt{Decl\to App}$.
Rule 29 instantiates $\al$ to the function type $\al[1] \to \al[2]$, just like Rules 10, 11 and 21.

@@

Figure~\ref{fig:alg:sample} shows a sample derivation of the algorithm.
It checks the application $(\lam x)~()$ against the unit type.
According to the algorithm, we apply Rule 18 (subsumption), changing to inference mode.
Type inference of the application breaks into two steps by Rule 26:
first we infer the type of the function,
and then the application inference judgment helps determine the return type.
In the following 5 steps The type of the identity function, $\lam x$, is inferred to be $\al \to \al$:
checking the body of the lambda function (Rule 25),
switching from check mode to inference mode (Rule 18),
inferring the type of a term variable (Rule 22),
solving a subtyping between existential variables (Rule 12) and
garbage collecting the term variable $x$ (Rule 3).

>>>>

The algorithmic typing reduction rules, defined in Figure~\ref{fig:alg}, have
the form $\Gm \rto \Gm'$.
% The worklists $\Gm, \Gm'$ contain both variable declarations and judgments.
The reduction process treats the worklist as a stack.  In every step it pops
the first judgment from the worklist for processing and possibly pushes new
judgments onto the worklist.  The syntax $\Gm\redto \Gm'$ denotes multiple
reduction steps. If $\Gm\redto\nil$, we say that $\Gm$ reduces successfully;
this corresponds to successful type checking.

Let us explain the rules in more detail:
Rules 1-3 pop a variable declaration. Essentially this is garbage: variables that are
out of scope for the remaining judgements in the worklist.
All other rules concern a judgement at the front of the worklist. Logically we
can discern 6 groups of rules.

@@

\paragraph{\bf 2. Existential decomposition.}
Rules 10 and 11 are algorithmic versions of Rule $\mathtt{{\le}{\to}}$; they
both partially instantiate $\al$ to function types.
The domain $\al[1]$ and range $\al[2]$ of the new function type are not determined:
they are fresh existential variables with the same scope as $\al$.
We replace $\al$ in the worklist with  $\al[1], \al[2]$.
To propagate the instantiation to the rest of the worklist and maintain well-formedness,
every reference to $\al$ is replaced by $\al[1] \to \al[2]$.
The \emph{occurs-check} condition prevents divergence as usual.
For example, without it $\al \le 1 \to \al$ would diverge.

@@

\paragraph{\bf 5. Inference judgments.}
Inference judgments behave differently compared with subtyping and checking judgments:
they \emph{return} a type instead of only accepting or rejecting.
For the algorithmic system, where guesses are involved,
it may happen that the output type of an inference judgment refers to new existential variables,
such as Rule 25.
In comparison to Rule 8 and 9, where new variables are only refered by the sub-derivation,
Rule 25 introduces variables $\al, \bt$ that affect the remaining judgment chain.
This rule is carefully designed so that the output variables are bound by earlier declarations,
thus the well-formedness of the worklist is preserved,
and the garbage will be collected at the correct time.
By making use of the continuation-passing-style judgement chain,
inner judgments always share the context with their parent judgment.

@@

Rule 25 infers the type of a lambda expression by introducing $\al, \bt$
as the input and output types of the function, respectively.
After checking the body $e$ under the assumption $x:\al$,
the return type might reflect more information than simply $\al \to \bt$
through propagation when existential variables are solved or partially solved.
The variable scopes are maintained during the process:
the assumption of argument type ($x:\al$) is recycled after checking against the function body;
the existential variables used by the function type ($\al,\bt$) are only visible in the remaining chain $[\al\to\bt/a]\jg$.
The recycling process of Rule 25 differs from DK's corresponding rule significantly,
and we further discuss the differences in Section~\ref{sec:discussion:scoping}.

@@

\paragraph{\bf 6. Application inference judgements}
Finally, Rules 27-29 deal with application inference judgments.
Rules 27 and 28 behaves similarly to declarative rules $\mathtt{Decl\forall App}$ and $\mathtt{Decl\to App}$.
Rule 29 instantiates $\al$ to the function type $\al[1] \to \al[2]$, just like Rules 10, 11 and 21.

@@

Figure~\ref{fig:alg:sample} shows a sample derivation of the algorithm.
It checks the application $(\lam x)~()$ against the unit type.
According to the algorithm, we apply Rule 18 (subsumption), changing to inference mode.
Type inference of the application breaks into two steps by Rule 26:
first we infer the type of the function,
and then the application inference judgment helps to determine the return type.
In the following 5 steps the type of the identity function, $\lam x$, is inferred to be $\al \to \al$:
checking the body of the lambda function (Rule 25),
switching from check mode to inference mode (Rule 18),
inferring the type of a term variable (Rule 22),
solving a subtyping between existential variables (Rule 12) and
garbage collecting the term variable $x$ (Rule 3).
